An Analysis of Health & Lifestyle Factors Associated with Sleep Quality

### 2024-05-03 products/manuscript/Manuscript.qmd
5/3- final touches

For our project, we obtained a dataset from Kaggle.com containing 373
observations of 13 variables from a study on different factors that may
affect sleep quality. Variables of interest from the study included
physical activity level, age, sex, occupation, stress, gender, BMI,
blood pressure, and sleep duration. In
[RStudio](https://posit.co/download/rstudio-desktop/) (version 4.3.1),
we used a combination of packages to create models suitable for
answering our primary question: which variables are most important for
determining an individual's sleep quality and what effect do they have?
After conducting an exploratory data analysis (EDA) we hypothesized that
stress, sleep duration, and occupation would have the largest impact on
sleep quality. We believed that stress and sleep quality would be
negatively correlated, sleep duration and quality would be positively
correlated, and that jobs with longer or more variable hours would
negatively affect sleep quality.

@@

The data was obtained from kaggle.com at this
[link](https://www.kaggle.com/datasets/henryshan/sleep-health-and-lifestyle/data).
The data contains observations from a study on individuals of varying
sexes and age regarding their sleep quality and different variables that
may affect this. We are unsure of where this data set came from and how
the data were collected; the publisher on Kaggle has yet to respond to a
comment with this question posted on March 5th 2024. Regardless, it
contains interesting variables that have been shown in previous studies
to have an effect on sleep quality.

For instance, in Sun et al. (2015), a sample of Chinese individuals
showed a correlation between obesity and worsened sleep quality in men
but not women. We are interested in seeing if this trend in present in
our dataset; however, given that most of the observations are of men, we
expect to see a correlation between lower sleep scores and obesity in
our data. Furthermore, it has been shown that age and gender have a
notable effect on sleep quality (Madrid-Valero et al. 2016). Women seem
to experience a deterioration in sleep quality as they age; however,
while this trend is still present in men, it is notably less consistent
and can vary dramatically between individuals. We expect to see similar
trends in our dataset, but as stated previously, our observations
contain significantly more men than women. Finally, we wish to test
whether or not certain occupational industries have different effects on
sleep quality. It has been shown that managerial positions tend to have
the worst sleep quality among civilian sector workers while 24 jobs that
have rotating shifts tend to have the worst sleep quality (Luckhaupt et
al. 2010). Our dataset contains observations of mostly
white-collar/service-based jobs, so we expect to find similar trends
upon analysis.

Additionally, based on the findings of Sun et. al that obesity can
negatively impact sleep quality, we also expect that higher levels of
blood pressure will result in poorer sleep quality. The American Sleep
Foundation states that both short and long sleep duration increase risk
of hypertension in adults (Li & Shang, 2021). This draws interesting
focus to the circular nature of sleep quality and hypertension. Those
with hypertension also face a higher risk of insomnia, a disorder
affecting the ability to fall asleep, maintain sleep, and sleep without
continuous disturbance (Liu et. al, 2022). Short sleep duration is also
positively associated with stress level (APA, 2013). Survey findings
indicate that the average American sleeps 6.7 hours per night, which is
well below the minimum recommended duration seven to nine hours (APA,
2013). Similar to the relationship between hypertension and sleep
quality, as stress level increases, sleep duration decreases thereby
causing poorer sleep quality. Poorer sleep quality then leads to higher
stress, greater risk of hypertension, and shorter sleep duration.

Engaging in physical activity has been suggested to improve both sleep
quality and sleep duration. Regular moderate to intense physical
activity stimulates an increase of melatonin production, a hormone that
regulates sleep-wake cycles (Alnawwar et. al, 2023). Additionally,
moderate to intense exercise has shown to have a positive effect on
those with insomnia (Alnawwar et. al, 2023). Physical activity also
reduces stress and improves both sleep quality and sleep efficiency
(Alnawwar et. al, 2023). However, studies also show that frequent
high-intensity workouts could lead to poorer sleep quality among people
with and without insomnia (Alnawwar et. al, 2023). Studies indicate that
consistent moderate exercise three to four times a week (approximately
150 minutes total) is the optimal weekly regimen to improve sleep
quality and reduce stress. We expect to see similar trends with our data
where increased stress levels negatively impact sleep quality. We expect
to see a positive association between increasing physical activity
levels and sleep quality until the point of "too much" activity.

@@

The dataset originally contained 373 observations for 13 variables
covering a broad spectrum of health and lifestyle variables associated
with sleep quality. The curators of the dataset did not provide
information on how or where this data was collected, so we solely relied
on the codebook for variable definitions. We have added this codebook to
the supplementary materials file for ease of access but you can also
find it on the Kaggle page for this dataset. There were no missing
values or erroneous variables in the dataset that we needed to remove.
The "Blood.Pressure" variable was transformed to reflect the individual
systolic and diastolic blood pressure measurements from each subject. We
then chose to add an additional variable, labeled 'cat_bp' to reflect
the categorical blood pressure status of a subject based on the
[American Heart Association
guidelines](https://www.heart.org/en/health-topics/high-blood-pressure).
Additionally we created a variable named 'StepsGroup' to determine the
categorical activity level of subjects in the dataset based on these
guidelines provided by [10,000
steps](https://www.10000steps.org.au/articles/healthy-lifestyles/counting-steps/#:~:text=The%20following%20pedometer%20indices%20have%20been%20developed%20to,day%205%20Highly%20active%20is%20more%20than%2012%2C500)
Finally, we created an additional categorical variable called
'PhysicalActivityGroup' to categorically represent the level of physical
activity reported by a subject. The levels are differentiated by 30
minute intervals of weekly physical activity. The outcome of interest is
noted as "Quality.of.Sleep" that we will reference as Sleep Quality or
Quality of Sleep throughout this report.

@@

The differences between the null model and RF model's metrics show that
our RF model is better suited for predicting sleep quality on unseen
data than a model using no predictors. The R-squared value of 0.997 is
extremely high; this shows that a large amount of variation is explained
using our RF model. The variable importance plot we created answers our
questions about which variables are most important for predicting sleep
quality. As we can see, sleep duration, stress level, and age were the
top three predictors for sleep quality in our data. This reflects the
relationships we saw in our colinearity plot as well as the clear
correlations we found between the variables in our EDA. We did not
expect to see age so high on the list; this could be due to us grouping
ages together in our EDA and therefore misleading us in our preliminary
data exploration. We did not use the groups when creating our models as
we learned that grouping continuous variables is a poor practice as we
progressed in our MADA course. Our RF model performed better than the LASSO 
model, which was to be expected. We struggled to create a variable importance
plot for the LASSO model, which could be due to sensitive parameters and 
tuning penalties.

>>>>

For our project, we obtained a dataset from Kaggle.com containing 373
observations of 13 variables from a study on different factors that may
affect sleep quality. Variables of interest from the study included
physical activity level, age, sex, occupation, stress, gender, BMI,
blood pressure, and sleep duration. In
[RStudio](https://posit.co/download/rstudio-desktop/) (version 4.3.1),
we used a combination of packages to create models suitable for
answering our primary question: which variables are most important for
determining an individual's sleep quality and what effect do they have [@noautho
r_rstudio_nodate]?
After conducting an exploratory data analysis (EDA) we hypothesized that
stress, sleep duration, and occupation would have the largest impact on
sleep quality. We believed that stress and sleep quality would be
negatively correlated, sleep duration and quality would be positively
correlated, and that jobs with longer or more variable hours would
negatively affect sleep quality.

@@

The data was obtained from kaggle.com at this
[link](https://www.kaggle.com/datasets/henryshan/sleep-health-and-lifestyle/data)
[@noauthor_find_nodate].
The data contains observations from a study on individuals of varying
sexes and age regarding their sleep quality and different variables that
may affect this. We are unsure of where this data set came from and how
the data were collected; the publisher on Kaggle has yet to respond to a
comment with this question posted on March 5th 2024. Regardless, it
contains interesting variables that have been shown in previous studies
to have an effect on sleep quality.

For instance, in Sun et al. (2015), a sample of Chinese individuals
showed a correlation between obesity and worsened sleep quality in men
but not women [@sun_poor_2016]. We are interested in seeing if this 
trend in present in
our dataset; however, given that most of the observations are of men, we
expect to see a correlation between lower sleep scores and obesity in
our data. Furthermore, it has been shown that age and gender have a
notable effect on sleep quality (Madrid-Valero et al. 2016)[@madrid-valero_age_
2017]. Women seem to experience a deterioration in sleep quality as they age; 
however,while this trend is still present in men, it is notably less consistent
and can vary dramatically between individuals. We expect to see similar
trends in our dataset, but as stated previously, our observations
contain significantly more men than women. Finally, we wish to test
whether or not certain occupational industries have different effects on
sleep quality. It has been shown that managerial positions tend to have
the worst sleep quality among civilian sector workers while 24 jobs that
have rotating shifts tend to have the worst sleep quality [@luckhaupt_prevalence
_2010]. Our dataset contains observations of mostly
white-collar/service-based jobs, so we expect to find similar trends
upon analysis.

Additionally, based on the findings of Sun et. al that obesity can
negatively impact sleep quality, we also expect that higher levels of
blood pressure will result in poorer sleep quality [@sun_poor_2016]. 
The American SleepFoundation states that both short and long sleep duration 
increase risk of hypertension in adults [@liu_association_2022]. This draws 
interesting focus to the circular nature of sleep quality and hypertension. 
Those with hypertension also face a higher risk of insomnia, a disorder
affecting the ability to fall asleep, maintain sleep, and sleep without
continuous disturbance [@liu_association_2022]. Short sleep duration is also
positively associated with stress level [@noauthor_stress_nodate]. Survey findings
indicate that the average American sleeps 6.7 hours per night, which is
well below the minimum recommended duration seven to nine hours (APA,
2013). Similar to the relationship between hypertension and sleep
quality, as stress level increases, sleep duration decreases thereby
causing poorer sleep quality. Poorer sleep quality then leads to higher
stress, greater risk of hypertension, and shorter sleep duration.

Engaging in physical activity has been suggested to improve both sleep
quality and sleep duration. Regular moderate to intense physical
activity stimulates an increase of melatonin production, a hormone that
regulates sleep-wake cycles [@alnawwar_effect_2023]. Additionally,
moderate to intense exercise has shown to have a positive effect on
those with insomnia [@alnawwar_effect_2023]. Physical activity also
reduces stress and improves both sleep quality and sleep efficiency
[@alnawwar_effect_2023]. However, studies also show that frequent
high-intensity workouts could lead to poorer sleep quality among people
with and without insomnia [@alnawwar_effect_2023]. Studies indicate that
consistent moderate exercise three to four times a week (approximately
150 minutes total) is the optimal weekly regimen to improve sleep
quality and reduce stress. We expect to see similar trends with our data
where increased stress levels negatively impact sleep quality. We expect
to see a positive association between increasing physical activity
levels and sleep quality until the point of "too much" activity.

@@

The dataset originally contained 373 observations for 13 variables
covering a broad spectrum of health and lifestyle variables associated
with sleep quality. The curators of the dataset did not provide
information on how or where this data was collected, so we solely relied
on the codebook for variable definitions. We have added this codebook to
the supplementary materials file for ease of access but you can also
find it on the Kaggle page for this dataset. There were no missing
values or erroneous variables in the dataset that we needed to remove.
The "Blood.Pressure" variable was transformed to reflect the individual
systolic and diastolic blood pressure measurements from each subject. We
then chose to add an additional variable, labeled 'cat_bp' to reflect
the categorical blood pressure status of a subject based on the
[American Heart Association
guidelines](https://www.heart.org/en/health-topics/high-blood-pressure) [@noauthor_understanding_nodate].
Additionally we created a variable named 'StepsGroup' to determine the
categorical activity level of subjects in the dataset based on these
guidelines provided by [10,000
steps](https://www.10000steps.org.au/articles/healthy-lifestyles/counting-steps
/#:~:text=The%20following%20pedometer%20indices%20have%20been%20developed%20to,
day%205%20Highly%20active%20is%20more%20than%2012%2C500) [@noauthor_counting_
nodate].
Finally, we created an additional categorical variable called
'PhysicalActivityGroup' to categorically represent the level of physical
activity reported by a subject. The levels are differentiated by 30
minute intervals of weekly physical activity. The outcome of interest is
noted as "Quality.of.Sleep" that we will reference as Sleep Quality or
Quality of Sleep throughout this report.

@@

The differences between the null model and RF model's metrics show that
our RF model is better suited for predicting sleep quality on unseen
data than a model using no predictors. The R-squared value of 0.997 is
extremely high; this shows that a large amount of variation is explained
using our RF model. The variable importance plot we created answers our
questions about which variables are most important for predicting sleep
quality. As we can see, sleep duration, stress level, and age were the
top three predictors for sleep quality in our data. This reflects the
relationships we saw in our collinearity plot as well as the clear
correlations we found between the variables in our EDA. We did not
expect to see age so high on the list; this could be due to us grouping
ages together in our EDA and therefore misleading us in our preliminary
data exploration. We did not use the groups when creating our models as
we learned that grouping continuous variables is a poor practice as we
progressed in our MADA course. Our RF model performed better than the LASSO 
model, which was to be expected. We struggled to create a variable importance
plot for the LASSO model, which could be due to sensitive parameters and 
tuning penalties.
